---
title: "A short report on the livestock_data dataset"
author: "Jean-Baka Domelevo Entfellner"
format: odt # or docx, pdf, odt, etc...
pdf-engine: pdflatex # or xelatex on Windows
papersize: a4
---

## Introduction

This is a brief report on the initial analysis of the `livestock_data` dataset, created during the R training workshop delivered in [Addis Ababa](https://en.wikipedia.org/wiki/Addis_Ababa) on 27 -- 29 August 2024.

## Loading the dataset

We load the dataset using the `readxl` package. After silently loading the `tidyverse` package, we load the first sheet of the Excel file:

```{r}
#| include: false
library(tidyverse)
```

```{r}
library(readxl)
dat <- read_xlsx("livestock_data.xlsx", sheet = 1)
glimpse(dat) # glimpse() from tibble is a bit like str()
```

In this dataset, we have `r nrow(dat)` observations and `r ncol(dat)` variables.

The statistical summary of the dataset is below.

```{r}
#| echo: false
summary(dat)
```

## A simple scatterplot

We want to represent the potential association between the stocking rate and the number of **tropical livestock units (TLU)** sold per hectare. In order to do this, we plot in @fig-scatterplot a scatterplot using `ggplot2`.

```{r}
#| label: fig-scatterplot
#| caption: "My first scatterplot with ggplot2"
#| warning: false
dat |> ggplot(aes(x=stocking_rate, y=TLUsold_ha)) + geom_point()
```

## And now, a linear model

We test the linear model having the TLU sold by hectare as the response (aka. *dependent variable*) and the stocking rate, high_pred_cattle and the cow_inmilk variable.

### Recoding NAs

But before do that, we replace all NAs in cow_inmilk with 0 values (because this is what should have appeared in the dataset).

```{r}
# we first replace some NAs with 0s
dat |> mutate(cow_inmilk = ifelse(is.na(cow_inmilk), 0, cow_inmilk)) -> dat

# alternatively, a slightly shorter version of the same:
dat |> mutate(cow_inmilk = replace_na(cow_inmilk, 0)) -> dat

# OR, same:
dat |> replace_na(list(cow_inmilk = 0)) -> dat
```

### And then the linear model

```{r}
# we then build the model:
lm1 <- lm(stocking_rate ~ TLUsold_ha + stocking_rate + cow_inmilk, data = dat)

# and we print its results with summary():
summary(lm1)
```

### Showcasing Latex 

Let's write the definition of the variance:

$var = \frac{\sum_{i = 1}^n (x_i - \bar{x})^2}{n-1}$

